---
title: "Assignment Zero: Practicing the basics of R"
subtitle: "Introduction To Chaos Applied To Systems, Processes And Products (ETSIDI, UPM)"
author: "Alfonso Allen-Perkins, Juan Carlos Bueno and Eduardo Faleiro"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
  html_document:
    theme: united
    toc: true
    toc_float: true
---

## Exercise 1: Setting up R and RStudio

1. Install **R** and **RStudio** (if not already installed).
2. Load the necessary libraries: 
   ```{r}
   library(ggplot2)
   ```
3. Create a simple **vector**, **matrix**, and **data frame** in R and display their structure.
   ```{r}
   my_vector <- c(1, 2, 3, 4, 5)
   my_matrix <- matrix(1:9, nrow=3)
   my_data_frame <- data.frame(Name = c("Alice", "Bob"), Age = c(25, 30))
   ```
4. Print each object and check their class using `class()`.

---

## Exercise 2: Creating functions in R

1. Define a function to compute a parabola:
   ```{r}
   parabola <- function(x) {
     return(x^2)
   }
   ```
2. Define a function for a straight line:
   ```{r}
   straight_line <- function(x, m=1, b=0) {
     return(m * x + b)
   }
   ```
3. Apply these functions to a sequence of values and store results.
4. Use **ggplot2** to plot both functions over the range [-10, 10].

```{r}
   x_values <- seq(-10, 10, length.out = 100)
  datafr_parabola <- data.frame(x = x_values, y = parabola(x_values))
  datafr_line <- data.frame(x = x_values, y = straight_line(x_values, m=1, b=0))
   
   # Parabola plot
   ggplot(data = datafr_parabola, aes(x = x, y = y)) +
     geom_line(color = "blue") +
     xlab("x") +
     ylab("Parabola: y = x^2") +
     ggtitle("Plot of a parabola") +
     theme_bw()
   
   # Straight line plot
   ggplot(data = datafr_line, aes(x = x, y = y)) +
     geom_line(color = "red") +
     xlab("x") +
     ylab("Straight Line: y = x") +
     ggtitle("Plot of a straight line") +
     theme_bw()
```

---

## Exercise 3: Using a For Loop

1. Implement a **for loop** to compute the orbit of the **logistic map**:
   
   $$ x_{n+1} = r x_n (1 - x_n) $$
   
   using an initial condition \( x_0 \) and a given value of \( r \).
   
2. Store the first **50 iterations** and plot **\( x_n \) vs. \( n \)** using `ggplot2`.
   ```{r}
   logistic_map <- function(r, x0, n) {
     x_values <- numeric(n)
     x_values[1] <- x0
     for (i in 2:n) {
       x_values[i] <- r * x_values[i-1] * (1 - x_values[i-1])
     }
     return(data.frame(n = 1:n, x = x_values))
   }
   
   # Example usage
   r_value <- 3.7
   x0_value <- 0.5
   n_iter <- 50
   
   orbit_data <- logistic_map(r_value, x0_value, n_iter)
   
   ggplot(orbit_data, aes(x = n, y = x)) +
     geom_line(color = "blue") +
     xlab("Iteration (n)") +
     ylab("x_n") +
     ggtitle("Logistic Map Orbit for r = 3.7, x0 = 0.5") +
     theme_bw()
   ```
3. Try different values of \( r \) (e.g., 2.5, 3.0, 3.5, 4.0) and observe the behavior.

---

## Exercise 4: Finding and visualizing roots of functions

### **Task 1: Finding roots of a polynomial**
1. Find all the roots of the polynomial \( x^2 - 1 \):
   ```{r}
   fixed_points <- polyroot(c(-1, 0, 1))
   print(fixed_points)
   ```
::: {.info}
**Note:** To learn how to use the polyroot function, type ```?polyroot``` in the console.
:::
   
2. Plot the polynomial function and highlight the roots:
   ```{r}
   x_values <- seq(-2, 2, length.out = 100)
   y_values <- x_values^2 - 1
   
   polynomial_curve_df <- data.frame(x = x_values, y = y_values)
   
   ggplot(data = polynomial_curve_df, aes(x = x, y = y)) +
     geom_line(color = "blue") +
     geom_point(data = data.frame(x = Re(fixed_points), y = 0), aes(x = x, y = y), color = "red", size = 3) +
     xlab("x") +
     ylab("y = x^2 - 1") +
     ggtitle("Roots of the Polynomial x^2 - 1") +
     theme_bw()
   ```

### **Task 2: Finding roots of a non-polynomial function**
1. Define the function \( f(x) = e^x - 10 \cos(x) \) and find a root in a given interval (for example, $[0, 2]$):
   ```{r}
   dx_dt <- function(x) {
     exp(x) - 10 * cos(x)
   }
   
   fixed_point_0 <- uniroot(dx_dt, interval = c(0, 2))$root
   print(fixed_point_0)
   ```
::: {.info}
**Note:** To learn how to use the polyroot function, type ```?uniroot``` in the console.
:::
   
2. Plot the function and highlight the root:
   ```{r}
   x_values <- seq(-2, 3, length.out = 100)
   y_values <- exp(x_values) - 10 * cos(x_values)
   
   non_polynomial_curve_df <- data.frame(x = x_values, y = y_values) 
   
   ggplot(data = non_polynomial_curve_df, aes(x = x, y = y)) +
      geom_line(color = "blue") +
      geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
      annotate("point", x = fixed_point_0, y = 0, color = "red", size = 3) +
      xlab("x") +
      ylab("y = e^x - 10*cos(x)") +
      ggtitle("Root of e^x - 10*cos(x) in [0, 2]") +
      theme_bw()
   ```
   
---

## Exercise 5: Reproducing figures from Strogatz’s book (avalaible in Moodle)

### **Task 1: Reproduce Figure 2.7.3**
- Create an **R script** that uses `ggplot2` to reproduce **Figure 2.7.3** from Strogatz’s book.
- Save the script as `figure_2_7_3.R` and generate the corresponding plot.

### **Task 2: Reproduce Figure 10.2.3**
- Create another script, `figure_10_2_3.R`, to replicate **Figure 10.2.3** from Strogatz’s book.
- Save the resulting figures as PNG or PDF files.
- Submit both **scripts and figures** via email.

---

::: {.info}
**Note:** Ensure that your plots use sufficient data points and appropriate labels for clarity.
:::