---
title: "2D-Discrete Dynamical Systems (Assignment Sheet 3)"
subtitle: "Introduction To Chaos Applied To Systems, Processes And Products (ETSIDI, UPM)"
author: "Alfonso Allen-Perkins, Juan Carlos Bueno and Eduardo Faleiro"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
---

## Introduction: Plotting 2D-orbits in R

Before solving the exercises, let's go through an example of how to plot the orbits of a 2D map using R. We will use the Gingerbreadman map, which is a simple example of a chaotic dynamical system that generates intricate, non-repeating patterns from a piecewise linear transformation.

The Gingerbreadman map is given by the equations:

\[
x_{n+1} = 1 - y_n + |x_n|
\]
\[
y_{n+1} = x_n
\]

To visualize orbits of this system in R, follow these steps:

1. **Load ggplot2 for visualization:**
```{r}
library(ggplot2)
```

2. **Define the function to generate the Gingerbreadman map:**

 ```{r}
gingerbreadman_map <- function(x0, y0, n) {
  x <- numeric(n)
  y <- numeric(n)
  x[1] <- x0
  y[1] <- y0
  
  for (i in 2:n) {
    x[i] <- 1 - y[i-1] + abs(x[i-1])
    y[i] <- x[i-1]
  }
  
  data.frame(x, y)
}
 ```
2. **Generate orbit with initial conditions:**
 ```{r}
data_gbm <- gingerbreadman_map(1.1, 6.1, 10000)
 ```
3. **Plot the orbit using `ggplot2`:**
 ```{r}
ggplot(data_gbm, aes(x = x, y = y)) +
  geom_point(color = 'blue', alpha = 0.5, size = .5) +
  labs(title = "Gingerbreadman map", x = "x", y = "y") +
  theme_bw()
 ```

You can modify the parameters to observe different behaviors of the above map using `ggplot2`.

## Exercise 1: The Lotka-Volterra discrete map
The Lotka-Volterra discrete map models predator-prey interactions in a discrete-time framework, capturing oscillatory population dynamics and potential chaotic behavior. The long-term behavior of the system can converge to fixed points, periodic cycles, or chaotic attractors, depending on the parameter values.

The discrete 2D Lotka-Volterra map is given by:

\[
x_n = \frac{\alpha x_{n-1} - \beta x_{n-1} y_{n-1}}{1 + \gamma x_{n-1}}
\]
\[
y_n = \frac{\delta y_{n-1} - \epsilon x_{n-1} y_{n-1}}{1 + \eta y_{n-1}}
\]

where:

- Case 1: \( \alpha = 170, \beta = 11, \gamma = 2.7, \delta = 50, \epsilon = 1.7, \eta = 7, x_1 = 7, y_1 = 5 \)
- Case 2: \( \alpha = 1.001, \beta = 0.03, \gamma = 0.6, \delta = 1.002, \epsilon = 1.7, \eta = 0.9, x_1 = 0.0002, y_1 = 0.0006 \)

Plot the solution (orbit) of the discrete dynamical system for both cases.

## Exercise 2:  The Henon map (1976)
The Henon map is a simplified model of chaotic dynamics in two dimensions, originally derived from a model of stellar motion in galaxies. It exhibits a well-known strange attractor, which demonstrates sensitive dependence on initial conditions, a key feature of chaotic systems. 

The Henon map is given by the equations:

\[
x_{n+1} = y_n + 1 - a x_n^2
\]
\[
y_{n+1} = b x_n
\]

where \( a \) and \( b \) are adjustable parameters. Set \( a = 1.4 \) and \( b = 0.3 \), then plot \( x_n \) vs \( y_n \).

## Exercise 3: The standard map (Chirikov 1971)
The Standard Map, also known as the Chirikov-Taylor Map, is a key model in Hamiltonian chaos, describing the transition from regular to chaotic motion in conservative dynamical systems. Depending on the parameter \( b \), the system can display stable islands, chaotic seas, or a mixture of both.

The Standard map is defined as follows:

\[
x_{n+1} = y_{n+1} + x_n
\]
\[
y_{n+1} = y_n + b \sin(x_n)
\]

- (a) Plot various orbits for \( b = 0 \).
- (b) Plot the phase portrait for \( b = 0.5 \).
- (c) Show the phase portrait for \( b = 1 \) and \( b = 2 \).

## Exercise 4: The Lozi map (1978)
The Lozi map is a piecewise linear version of the Henon map that retains its chaotic properties while being computationally simpler to analyze. It generates a strange attractor with a fractal structure, making it an important example in the study of deterministic chaos.

The Lozi map is defined by:

\[
x_{n+1} = y_n + 1 - a |x_n|
\]
\[
y_{n+1} = b x_n
\]

where \( -1 < b < 1 \). Plot orbits when \( a = 1.7 \) and \( b = 0.5 \).
